---
title: "Land Cover Malawi"
author: "Nicholas & Victor"
date: "2026-04-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages

```{r}
#install.packages("terra")
#install.packages("geodata")
library(ggplot2)
library(gganimate)
library(gifski)
library(av)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(viridis)
library(terra)
library(geodata)
library(sf)
library(tidyverse)
```

Downloading Land Cover data from WorldCover using geodata package

```{r}

# names of the land cover types
vars <- c(
  "trees", "grassland", "shrubs", "cropland",
  "built", "bare",
  "wetland"
)

lc_list <- list() # empty list

# Download each dataset based on the list
for (v in vars) {
  lc_list[[v]] <- geodata::landcover(
    var  = v,
    path = "data/"
  )
}

```

Crop and mask the WorldCover data to Malawi region

```{r}
# Crop the dataset based on Malawi shapfile
mwi <- geodata::gadm(country = "MWI", level = 0, path = "data/")

lc_mwi <- lapply(lc_list, function(r) {
  terra::mask(terra::crop(r, mwi), mwi)
})

```

Stack the rasters and plot 

```{r}
lc_stack <- rast(lc_mwi)
plot(lc_stack)
```
Get Malawi Districts shapefiles and set CRS

```{r}
#Geolocating points in the climate data and linking it to distrits for merging with the malaria data

# Load Malawi districts shapefile

districts <- st_read("mwi_adm_nso_hotosm_20230405_shp/mwi_admbnda_adm2_nso_hotosm_20230405.shp")

# Transform to same CRS as your climate points
districts <- st_transform(districts, crs = 4326)  # WGS84

```

Plot Malawi map with district labels

```{r}
districts_sf <- st_as_sf(districts)

# Plot
ggplot() +
  geom_sf(data = districts_sf, fill = NA, color = "black") +
  geom_sf_text(data = districts_sf, aes(label = ADM2_EN), size = 2, color = "blue") +
  theme_minimal() +
  labs(title = "Malawi Districts", caption = "Source: Malawi shapefile")

```


```{r}

# Get mean fraction of land cover of each District for analysis
# Later we may get more data

districts_vect <- vect(districts) # making sure it is a spatial vector
districts_vect <- project(districts_vect, crs(lc_stack))

lc_by_district <- terra::extract(
  lc_stack,
  districts_vect,
  fun = mean,
  na.rm = TRUE
)

# Prepare the mean land cover fraction dataset for analysis 
lc_by_district$District <- districts_vect$ADM2_EN
lc_df <- lc_by_district[, c("District", vars)]

head(lc_df)
```
Plot maps of mean land cover fraction by each District for easier visualization

```{r}


# Join lc_df (mean land cover) to districts_sf
districts_sf <- districts_sf %>%
  left_join(lc_df, by = c("ADM2_EN" = "District"))

# Plot maps in a loop
vars <- c(
  "trees", "grassland", "shrubs", "cropland",
  "built", "bare",
  "wetland"
)

plot_list <- list()

for (v in vars) {
  
  p <- ggplot(districts_sf) +
    geom_sf(aes_string(fill = v), color = "black") +
    scale_fill_viridis_c(option = "plasma", name = paste("%", v)) +
    theme_minimal() +
    labs(
      title = paste("Mean", v, "fraction by District in Malawi"),
      caption = "Source: ESA WorldCover"
    )
  
  plot_list[[v]] <- p
  print(p)
  
  # save each map
  # ggsave(filename = paste0(v, "_malawi_map.png"), plot = p, width = 8, height = 6)
}

```



```{r}

# Filter Dowa district only
dowa_sf <- districts_sf %>%
  filter(ADM2_EN == "Dowa")

# Convert to terra vector
dowa_vect <- vect(dowa_sf)

# Project Dowa polygon to match landcover raster CRS
dowa_vect <- project(dowa_vect, crs(lc_stack))

plot(dowa_vect)
# Extract mean landcover values for Dowa only
lc_dowa_mean <- terra::extract(
  lc_stack,
  dowa_vect,
  fun = mean,
  na.rm = TRUE
)

# Add district name
lc_dowa_mean$District <- dowa_vect$ADM2_EN

# Keep only useful columns
lc_dowa_mean <- lc_dowa_mean[, c("District", vars)]

lc_dowa_mean

```

Population Density data for Dowa at 1 km-square grid

```{r}
library(terra)
library(sf)
library(dplyr)
library(geodata)


# Download population density raster
# res can be 0.5, 2.5, 5, or 10 minutes depending on what is available
pop_density <- geodata::population(
  year = 2020,
  res  = 0.5,
  path = "data/"
)

# Search for downloaded population raster files
pop_files <- list.files(
  "data/",
  recursive = TRUE,
  full.names = TRUE,
  pattern = "\\.tif$"
)

pop_density <- rast(pop_files[8])

terra::hasValues(pop_density)

plot(pop_density, main = "Population density")

```

Extract for Dowa
```{r}

# Project Dowa polygon to population raster CRS
dowa_vect_pop <- project(dowa_vect, crs(pop_density))

# Crop and mask population density raster to Dowa
pop_density_dowa <- pop_density %>%
  terra::crop(dowa_vect_pop) %>%
  terra::mask(dowa_vect_pop)

names(pop_density_dowa) <- "pop_density_per_km2"

plot(pop_density_dowa, main = "Population density in Dowa")
plot(dowa_vect_pop, add = TRUE, border = "red", lwd = 2)
```

Extract mean population density for Dowa
```{r}
mean_pop_density_dowa <- terra::extract(
  pop_density_dowa,
  dowa_vect,
  fun = mean,
  na.rm = TRUE
)

mean_pop_density_dowa$District <- "Dowa"

mean_pop_density_dowa
```
Join population density to land-cover points

```{r}
mean_pop_density_dowa <- mean_pop_density_dowa %>%
  select(District, pop_density_per_km2)

dowa_average_lc_pop <- lc_dowa_mean %>%
  left_join(mean_pop_density_dowa, by = "District")

dowa_average_lc_pop
```

Malaria data for Dowa for each Month since 2018
```{r}
incidence_data <- read_csv("malaria.csv") %>% 
  filter(District =="Dowa") %>% 
  group_by(Month, Year) %>%
  summarise(Incidence = sum(Incidence, na.rm = TRUE), .groups = "drop") %>% 
  mutate( Month = factor(Month, levels = month.name, ordered = TRUE) ) %>%
  arrange(Year, Month)

ggplot(incidence_data, aes(x = Month, y = Incidence, group = Year, color = factor(Year))) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Incidence",
    color = "Year"
  )
```
Create a detailed pointwise location for Dowa: Start from your Dowa land-cover points
```{r}
# Convert Dowa polygon to terra vector
dowa_vect <- terra::vect(dowa_sf)

# Match CRS with land-cover raster
dowa_vect_lc <- terra::project(dowa_vect, terra::crs(lc_stack))

# Crop and mask land cover to Dowa
lc_dowa <- lc_stack %>%
  terra::crop(dowa_vect_lc) %>%
  terra::mask(dowa_vect_lc)

# Convert raster cells to dataframe with coordinates
lc_dowa_points <- as.data.frame(
  lc_dowa,
  xy = TRUE,
  na.rm = TRUE
) %>%
  rename(
    longitude = x,
    latitude  = y
  ) %>%
  mutate(District = "Dowa") %>%
  select(District, longitude, latitude, all_of(vars))
```

Now, create  Create dowa_lc_pop which is a file that extracts population density at each land-cover raster-cell coordinate.
```{r}
# Convert land-cover point dataframe to terra vector
lc_dowa_pts_vect <- terra::vect(
  lc_dowa_points,
  geom = c("longitude", "latitude"),
  crs  = terra::crs(lc_stack)
)

# Project land-cover points to population raster CRS
lc_dowa_pts_vect_pop <- terra::project(
  lc_dowa_pts_vect,
  terra::crs(pop_density_dowa)
)

# Extract population density at each land-cover point
pop_at_lc_points <- terra::extract(
  pop_density_dowa,
  lc_dowa_pts_vect_pop
)

# Create merged detailed dataset
dowa_lc_pop <- lc_dowa_points %>%
  mutate(
    pop_density_per_km2 = pop_at_lc_points$pop_density_per_km2
  )

head(dowa_lc_pop)
```

Now, I use dowa_lc_pop as the sampling frame for the weighted malaria case allocation.

Now, I define dominant land cover for each raster cell

Because each raster cell may have fractions of trees, grassland, shrubs, cropland, built, bare, and wetland, I assign the dominant land-cover class as the one with the highest fraction.
```{r}
landcover_vars <- c(
  "trees", "grassland", "shrubs", "cropland",
  "built", "bare", "wetland"
)

dowa_sampling_frame <- dowa_lc_pop %>%
  rowwise() %>%
  mutate(
    dominant_landcover = landcover_vars[
      which.max(c_across(all_of(landcover_vars)))
    ]
  ) %>%
  ungroup()

## Check its distribution
dowa_sampling_frame %>%
  count(dominant_landcover) %>%
  mutate(prop = n / sum(n))
```

I now create a population, urban, and land-cover weights where: 

*pop_weight: gives higher probability to cells with higher population density.
*urban_weight: gives extra probability to cells with more built-up land cover.
*landcover_weight: gives extra probability to land-cover types that may be more relevant for malaria exposure.

```{r}
dowa_sampling_frame <- dowa_sampling_frame %>%
  mutate(
    # Avoid zero population weights
    pop_density_clean = ifelse(
      is.na(pop_density_per_km2),
      0,
      pop_density_per_km2
    ),
    
    pop_weight = pop_density_clean + 1,
    
    # Urban/built-up weight
    urban_weight = 1 + built,
    
    # Land-cover suitability weights based on Doreen's landcover against malaria incidence results
    landcover_weight = case_when(
      dominant_landcover == "cropland"  ~ 1.50,
      dominant_landcover == "grassland" ~ 1.20,
      dominant_landcover == "shrubs"    ~ 1.10,
      dominant_landcover == "wetland"   ~ 1.00,
      dominant_landcover == "built"     ~ 1.30,
      dominant_landcover == "trees"     ~ 0.80,
      dominant_landcover == "bare"      ~ 0.60,
      TRUE                              ~ 1.00
    )
  )
```

I am now creating seasonal weights from the incidence pattern. This is a month-specific seasonal multiplier based on the average incidence across years.
```{r}
seasonal_weights <- incidence_data %>%
  group_by(Month) %>%
  summarise(
    mean_incidence = mean(Incidence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    seasonal_weight = mean_incidence / mean(mean_incidence, na.rm = TRUE)
  )

seasonal_weights
```

I then create 70/30 sampling structure by land cover.Here I define two “high-priority” land-cover classes, cropland and grassland to have 70% of cases and 30% from all other land-cover types.I then create a
sampling function for one month, that takes one row of incidence data and returns sampled locations. The final sampling weight= population weight * urban/built weight * land-cover suitability weight * seasonal month weight
```{r}
sample_dowa_cases_one_month <- function(inc_row,
                                        sampling_frame,
                                        seasonal_weights,
                                        priority_landcovers = c("cropland", "grassland"),
                                        priority_share = 0.70,
                                        other_share = 0.30,
                                        seed = NULL) {
  
  if (!is.null(seed)) set.seed(seed)
  
  this_month <- inc_row$Month
  this_year  <- inc_row$Year
  n_cases    <- inc_row$Incidence
  
  # If no cases, return empty data frame
  if (is.na(n_cases) || n_cases <= 0) {
    return(data.frame())
  }
  
  # Get seasonal weight for the month
  sw <- seasonal_weights %>%
    filter(Month == this_month) %>%
    pull(seasonal_weight)
  
  if (length(sw) == 0 || is.na(sw)) sw <- 1
  
  # Add final sampling weights
  sf <- sampling_frame %>%
    mutate(
      month = as.character(this_month),
      year = this_year,
      seasonal_weight = sw,
      final_weight = pop_weight *
        urban_weight *
        landcover_weight *
        seasonal_weight
    )
  
  # Split into priority and non-priority land-cover cells
  sf_priority <- sf %>%
    filter(dominant_landcover %in% priority_landcovers)
  
  sf_other <- sf %>%
    filter(!dominant_landcover %in% priority_landcovers)
  
  # Number of samples from each group
  n_priority <- round(n_cases * priority_share)
  n_other    <- n_cases - n_priority
  
  # Safety checks
  if (nrow(sf_priority) == 0) {
    sf_priority <- sf
    n_priority <- n_cases
    n_other <- 0
  }
  
  if (nrow(sf_other) == 0) {
    sf_other <- sf
    n_other <- 0
  }
  
  # Sample priority cells
  sampled_priority <- sf_priority %>%
    slice_sample(
      n = n_priority,
      replace = TRUE,
      weight_by = final_weight
    )
  
  # Sample other cells
  sampled_other <- sf_other %>%
    slice_sample(
      n = n_other,
      replace = TRUE,
      weight_by = final_weight
    )
  
  # Combine
  sampled <- bind_rows(sampled_priority, sampled_other) %>%
    mutate(
      sampled_group = ifelse(
        dominant_landcover %in% priority_landcovers,
        "priority_landcover",
        "other_landcover"
      )
    )
  
  sampled
}
```


Now, run sampling for all months and years. Because my monthly incidence can be large, this may create many rows. For example, if Dowa has 50,000 cases in a month, the output will contain 50,000 sampled rows for that month.
```{r}
priority_landcovers <- c("cropland", "wetland")
sampled_dowa_cases <- lapply(seq_len(nrow(incidence_data)), function(i) {
  
  sample_dowa_cases_one_month(
    inc_row = incidence_data[i, ],
    sampling_frame = dowa_sampling_frame,
    seasonal_weights = seasonal_weights,
    priority_landcovers = priority_landcovers,
    priority_share = 0.70,
    other_share = 0.30,
    seed = 100 + i
  )
  
}) %>%
  bind_rows()
```

Check: has the weighting being done corectly 
```{r}

sampled_dowa_cases %>%
  count(year, month, dominant_landcover)

##Has 70/30 worked?
sampled_dowa_cases %>%
  count(year, month, sampled_group) %>%
  group_by(year, month) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()
```

Now, I keeps only the number of sampled cases per raster cell which we will be using for modelling 
```{r}
sampled_dowa_cases_aggregated <- sampled_dowa_cases %>%
  group_by(
    year,
    month,
    District,
    longitude,
    latitude,
    dominant_landcover
  ) %>%
  summarise(
    sampled_cases = n(),
    pop_density_per_km2 = first(pop_density_per_km2),
    trees = first(trees),
    grassland = first(grassland),
    shrubs = first(shrubs),
    cropland = first(cropland),
    built = first(built),
    bare = first(bare),
    wetland = first(wetland),
    .groups = "drop"
  )

head(sampled_dowa_cases_aggregated)
```

```{r}


library(tidyverse)

dir.create("monthly_maps", showWarnings = FALSE)

sampled_dowa_cases_aggregated <- sampled_dowa_cases_aggregated %>%
  mutate(
    month = factor(month, levels = month.name),
    year_month = paste(year, month, sep = "_")
  )

year_month_list <- sampled_dowa_cases_aggregated %>%
  distinct(year, month, year_month) %>%
  arrange(year, month)

for(i in seq_len(nrow(year_month_list))) {
  
  y <- year_month_list$year[i]
  m <- year_month_list$month[i]
  
  p <- sampled_dowa_cases_aggregated %>%
    filter(year == y, month == m) %>%
    ggplot(aes(x = longitude, y = latitude)) +
    geom_point(
      aes(color = sampled_cases),
      size = 1,
      alpha = 0.8
    ) +
    scale_color_gradientn(
      colors = c("lightyellow", "orange", "red", "darkred"),
      limits = c(0, 30),
      oob = scales::squish
    ) +
    coord_equal() +
    theme_minimal() +
    labs(
      title = paste("Sampled malaria cases -", m, y),
      x = "Longitude",
      y = "Latitude",
      color = "Cases"
    )
  
  ggsave(
    filename = paste0("monthly_maps/map_", y, "_", m, ".jpg"),
    plot = p,
    width = 7,
    height = 6,
    dpi = 300
  )
}
```


# include more shapefiles form Malawi later: https://data.humdata.org/dataset/cod-ab-mwi

```{r}
library(magick)

png_files <- sort(list.files(
  "monthly_maps",
  pattern = "\\.jpg$",
  full.names = TRUE
))

imgs <- image_read(png_files)

gif <- image_animate(imgs, fps = 2)

image_write(gif, "monthly_malaria_cases.gif")
```
