---
title: "Climatic Based Malaria Incidence Prediction in Malawi"
author:
  name: Doreen A. Ndovie, Dr Sureni Wickramasooria, Nicholas S. Adam PhD
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
link-citations: true
execute:
  echo: true
  warning: false
  message: false
---

## Project Description

Malaria remains one of the most persistent public health challenges in Malawi. Every year, thousands of people — especially children and pregnant women — are at risk of this deadly disease. This analysis aims to support predictive modelling of malaria outbreaks using ERA5 climate data, land cover classifications, and district-level malaria incidence records from 2015 to 2024. Specific objectives include:

- Assessing and transforming climate and land cover data as required
- Correlating land cover characteristics with total malaria incidence per district
- Correlating ERA5 climate variables with malaria incidence using 1-, 2-, and 3-month lags
- Exporting a clean, analysis-ready dataset with lagged climate features

---

# Libraries

```{r libraries}
library(tidyverse)
library(lubridate)
library(broom)
library(stringr)
library(stringi)
library(scales)
library(patchwork)
library(tseries)
library(forecast)
library(kableExtra)
library(readxl)
library(purrr)
library(imputeTS)
library(data.table)
library(writexl)
library(sf)
library(corrr)
library(corrplot)
library(Hmisc)
library(car)
library(grid)
library(terra)
library(ggrepel)
library(openxlsx)
library(ggspatial)
```

```{r r-citation, echo=FALSE}
# R citation for reproducibility
cat("R version citation:\n")
citation()
cat("\nR version in use:\n")
R.version.string
```

---

## ERA5 Climate Data

The ERA5 dataset is already aggregated at district level and contains: `district`, `date`, `temperature`, `precipitation`, and `wind`. No spatial geolocating is required.

**Data source:** ERA5 hourly data on single levels from 1940 to present. Copernicus Climate Change Service (C3S) Climate Data Store. Hersbach, H. et al. (2020). The ERA5 global reanalysis. *Quarterly Journal of the Royal Meteorological Society*, 146, 1999–2049. doi:10.1002/qj.3803. Accessed via the Copernicus Climate Data Store (CDS): https://cds.climate.copernicus.eu/. Variables used: 2-metre air temperature (°C), total precipitation (mm), and 10-metre wind speed (m/s), aggregated to district level using area-weighted spatial averaging over Malawi administrative boundaries (2018–2024).

```{r import-climate}
climate_raw <- read_csv("Malawi_district_climate_2015_2024.csv")

glimpse(climate_raw)
```

## Land Cover Data

```{r import-landcover}
landcover_df <- read_csv("landcover.csv")

glimpse(landcover_df)
head(landcover_df)
```

## Malaria Incidence Data

The malaria incidence data is read from `malaria_incidence_cleaned.csv`, which contains district-level monthly case counts with columns: `District`, `Month`, `Year`, `Cases`.

```{r import-malaria}
malaria_raw <- read_csv("malaria_incidence_cleaned.csv")

# Standardise column names to lowercase
names(malaria_raw) <- tolower(names(malaria_raw))

# Rename 'cases' to 'incidence' for consistency with downstream code
malaria_raw <- malaria_raw %>%
  rename(incidence = cases)

glimpse(malaria_raw)
head(malaria_raw)
```

## Population Data (NSO - Cleaning)

```{r import-population}
pop_dat <- read_csv("MR DATA - District Populations.csv")

pop_tot <- pop_dat %>%
  rename(AgeGroup = 2) %>%
  filter(AgeGroup == "Total") %>%
  filter(!is.na(District))

# Rename the repeating triplets: 2018_both, 2018_male, 2018_female
yrs <- 2015:2024
nm  <- names(pop_tot)

year_pos <- which(nm %in% as.character(yrs))

for (y in yrs) {
  i <- which(nm == as.character(y))
  if (length(i) == 1) {
    nm[i]     <- paste0(y, "_both")
    if (i + 1 <= length(nm)) nm[i + 1] <- paste0(y, "_male")
    if (i + 2 <= length(nm)) nm[i + 2] <- paste0(y, "_female")
  }
}
names(pop_tot) <- nm

wanted <- c("District",
            as.vector(rbind(paste0(yrs, "_both"),
                            paste0(yrs, "_male"),
                            paste0(yrs, "_female"))))
wanted <- intersect(wanted, names(pop_tot))

pop_long <- pop_tot %>%
  select(all_of(wanted)) %>%
  pivot_longer(
    cols      = -District,
    names_to  = c("Year", "Sex"),
    names_sep = "_",
    values_to = "Population"
  ) %>%
  mutate(
    Year       = as.integer(Year),
    Sex        = factor(Sex, levels = c("both", "male", "female"),
                        labels = c("Both", "Male", "Female")),
    Population = parse_number(Population)
  ) %>%
  arrange(District, Year, Sex)

pop_long_all <- pop_long %>% filter(Year %in% 2015:2024)

pop_both_combined <- pop_long_all %>%
  mutate(
    District_clean = str_replace(District, "\\s+(City|Rural)$", ""),
    District_clean = str_squish(District_clean),
    District_clean = case_when(
      District_clean == "Chiradzulo" ~ "Chiradzulu",
      TRUE ~ District_clean
    )
  ) %>%
  filter(Sex == "Both") %>%
  group_by(District_clean, Year) %>%
  summarise(Population = sum(Population, na.rm = TRUE), .groups = "drop") %>%
  arrange(District_clean, Year) %>%
  mutate(District_clean = District_clean %>%
           stringr::str_replace_all("[^[:alnum:] ]", "") %>%
           stringr::str_squish())

# The source population data may not cover every year in 2015-2024 (e.g. only
# 2018-2024 may be present). To avoid NA/NaN incidence_per_1k for the missing
# years, build a complete District x Year grid (2015-2024) and carry the
# nearest available year's population forward/backward as an estimate.
pop_grid <- expand_grid(
  District_clean = unique(pop_both_combined$District_clean),
  Year           = 2015:2024
)

pop_both_combined <- pop_grid %>%
  left_join(pop_both_combined, by = c("District_clean", "Year")) %>%
  arrange(District_clean, Year) %>%
  group_by(District_clean) %>%
  fill(Population, .direction = "downup") %>%
  ungroup()

cat("District/year combinations still missing population after fill:\n")
pop_both_combined %>% filter(is.na(Population)) %>% print(n = Inf)
```

---

# Climate Data Cleaning

## Filter to Study Period (2018–2024) and Aggregate Monthly

```{r climate-filter}
climate_filtered <- climate_raw %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= as.Date("2018-01-01"),
         date <= as.Date("2024-12-31"))

climate_monthly <- climate_filtered %>%
  mutate(
    year  = lubridate::year(date),
    month = lubridate::month(date, label = TRUE, abbr = TRUE)
  ) %>%
  group_by(district, year, month) %>%
  summarise(
    wind          = mean(wind,         na.rm = TRUE),
    temperature   = mean(temperature,  na.rm = TRUE),
    precipitation = sum(precipitation, na.rm = TRUE),
    .groups = "drop"
  )

head(climate_monthly)
```

## Final Climate Data Preparation

Standardise district names, convert abbreviated month labels to full names, and construct a proper monthly date column.

```{r climate-clean-final}
climate_df <- climate_monthly %>%
  mutate(
    district = case_when(
      tolower(district) %in% c("lilongwe city", "lilongwe city council") ~ "Lilongwe",
      tolower(district) == "nkhata bay" ~ "Nkhatabay",
      TRUE ~ district
    ),
    month = factor(
      month,
      levels = month.abb,
      labels = month.name
    ),
    month_num = match(as.character(month), month.name),
    date      = as.Date(paste(year, month_num, "01", sep = "-"))
  )

head(climate_df)
unique(climate_df$district)
```

## Check Distributions and Apply Transformations

```{r climate-transform}
climate_df %>%
  select(precipitation, temperature, wind) %>%
  summary()

hist(
  climate_df$precipitation,
  main   = "Precipitation Distribution (raw)",
  xlab   = "Precipitation (mm)",
  col    = "steelblue",
  breaks = 30
)

climate_df <- climate_df %>%
  mutate(log_precipitation = log1p(precipitation))

hist(
  climate_df$log_precipitation,
  main   = "Precipitation Distribution (log1p transformed)",
  xlab   = "log(Precipitation + 1)",
  col    = "steelblue",
  breaks = 30
)

hist(
  climate_df$temperature,
  main   = "Temperature Distribution",
  xlab   = "Temperature (°C)",
  col    = "tomato",
  breaks = 30
)

hist(
  climate_df$wind,
  main   = "Wind Speed Distribution",
  xlab   = "Wind Speed (m/s)",
  col    = "darkseagreen",
  breaks = 30
)
```

## Climate Trends per District (2015–2024)

```{r climate-trend-plots, fig.width=12, fig.height=16}
climate_long <- climate_df %>%
  select(district, date, precipitation, temperature, wind) %>%
  pivot_longer(
    cols      = c(precipitation, temperature, wind),
    names_to  = "variable",
    values_to = "value"
  )

full_grid <- expand.grid(
  date     = seq(as.Date("2018-01-01"), as.Date("2024-12-01"), by = "month"),
  district = unique(climate_long$district),
  stringsAsFactors = FALSE
)

climate_long <- full_grid %>%
  left_join(climate_long, by = c("date", "district"))

year_start_dates <- unique(climate_long$date[format(climate_long$date, "%m") == "01"])

base_dir <- "climate_plots_by_district"
dir.create(base_dir, showWarnings = FALSE)

for (d in unique(climate_long$district)) {
  district_dir  <- file.path(base_dir, d)
  dir.create(district_dir, showWarnings = FALSE)
  district_data <- filter(climate_long, district == d)

  p <- ggplot(district_data, aes(x = date, y = value)) +
    geom_smooth(method = "loess", span = 0.25, se = FALSE,
                color = "steelblue", linewidth = 1.1, na.rm = TRUE) +
    geom_vline(xintercept = as.numeric(year_start_dates),
               color = "grey70", linetype = "dashed", linewidth = 0.5) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month",
                 limits = c(as.Date("2018-01-01"), as.Date("2024-12-01"))) +
    facet_wrap(~variable, scales = "free_y", ncol = 1) +
    theme_minimal() +
    theme(
      axis.text.x      = element_text(angle = 60, hjust = 1, size = 7),
      strip.text       = element_text(face = "bold", size = 12),
      panel.spacing    = unit(1.6, "lines"),
      panel.grid.minor = element_blank()
    ) +
    labs(title = NULL, x = NULL, y = NULL)

  ggsave(
    filename = file.path(district_dir, paste0("climate_trends_", d, ".pdf")),
    plot = p, width = 12, height = 16, units = "in", device = "pdf"
  )
}

message("Climate trend plots saved to: ", base_dir)
```

---

# Landcover Data Cleaning

```{r landcover-clean}
landcover_df <- landcover_df %>%
  rename_with(tolower) %>%
  mutate(district = str_to_title(str_trim(district)))

cat("Missing values per column:\n")
print(colSums(is.na(landcover_df)))

landcover_vars <- landcover_df %>%
  select(where(is.numeric)) %>%
  names()

cat("\nLand cover variables detected:\n")
print(landcover_vars)

landcover_df %>%
  select(all_of(landcover_vars)) %>%
  summary()

par(mfrow = c(ceiling(length(landcover_vars) / 3), 3), mar = c(4, 4, 2, 1))
for (v in landcover_vars) {
  hist(landcover_df[[v]], main = v, xlab = v, col = "darkseagreen", breaks = 15)
}
par(mfrow = c(1, 1))

cat("\nLand cover districts:\n")
print(unique(landcover_df$district))
```

---

# Malaria Data Cleaning

## Harmonise District Names and Clean

```{r malaria-clean}
malaria_df <- malaria_raw %>%
  # Step 1: strip -DHO and normalise whitespace/case for safe matching
  mutate(
    district_raw = district %>%
      stringr::str_remove("-DHO$") %>%
      stringr::str_squish(),
    district = case_when(
      str_detect(tolower(district_raw), "nkhata")   ~ "Nkhatabay",
      str_detect(tolower(district_raw), "mzimba")   ~ "Mzimba",   # catches North, South, plain
      str_detect(tolower(district_raw), "mzuzu")    ~ "Mzimba",
      TRUE ~ district_raw %>%
               stringr::str_remove_all("[^[:alnum:] ]") %>%
               stringr::str_squish()
    )
  ) %>%
  select(-district_raw) %>%
  mutate(
    year  = as.integer(year),
    month = factor(month, levels = c("January","February","March","April",
                                     "May","June","July","August","September",
                                     "October","November","December"))
  ) %>%
  filter(year >= 2015, year <= 2024)

cat("Unique districts after cleaning:\n")
print(sort(unique(malaria_df$district)))
```

## Attach Population and Compute Incidence per 1,000

```{r malaria-pop}
malaria_with_pop <- malaria_df %>%
  left_join(
    pop_both_combined,
    by = c("district" = "District_clean", "year" = "Year")
  ) %>%
  mutate(incidence_per_1k = (incidence / Population) * 1000)

glimpse(malaria_with_pop)
```

## Create Date Column

```{r malaria-date}
malaria_with_pop <- malaria_with_pop %>%
  mutate(
    month_num = match(tolower(trimws(as.character(month))), tolower(month.name)),
    date      = as.Date(paste(year, month_num, "01", sep = "-"))
  ) %>%
  filter(year >= 2015 & year <= 2024)
```

---

# Data Merging

## Aggregate Malaria to District-Month Level

```{r merge-aggregate}
district_monthly <- malaria_with_pop %>%
  group_by(district, year, month, date) %>%
  summarise(
    total_cases      = sum(incidence,   na.rm = TRUE),
    total_population = mean(Population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # mean() of an all-NA group with na.rm = TRUE returns NaN, not NA, which
    # would otherwise propagate into incidence_per_1k. Guard against that
    # (and against division by zero) here.
    total_population = ifelse(is.nan(total_population) | total_population == 0,
                               NA_real_, total_population),
    incidence_per_1k  = (total_cases / total_population) * 1000
  )

cat("District-month rows with missing population (and thus NA incidence_per_1k):\n")
district_monthly %>% filter(is.na(total_population)) %>% distinct(district, year)

head(district_monthly)
```

## Merge Climate and Malaria Data

```{r merge-climate-malaria}
climate_df$month       <- as.character(climate_df$month)
district_monthly$month <- as.character(district_monthly$month)

merged_df <- district_monthly %>%
  # ERA5 climate data is only available for 2018-2024, so restrict the
  # climate-malaria merge to that period. district_monthly (2015-2024)
  # remains unchanged for incidence-only visualisations.
  filter(year >= 2018) %>%
  left_join(
    climate_df %>% mutate(year = as.numeric(as.character(year))),
    by = c("district", "date", "year", "month")
  )

head(merged_df)

cat("Rows in merged_df:", nrow(merged_df), "\n")
cat("NAs in precipitation:", sum(is.na(merged_df$precipitation)), "\n")
cat("NAs in temperature:",   sum(is.na(merged_df$temperature)),   "\n")
cat("NAs in wind:",          sum(is.na(merged_df$wind)),          "\n")
```

## Merge Land Cover Data

```{r merge-landcover}
# Land cover vs incidence correlations are based on 2020 incidence data only
district_total_incidence <- district_monthly %>%
  filter(year == 2020) %>%
  group_by(district) %>%
  summarise(
    total_incidence  = sum(total_cases,       na.rm = TRUE),
    avg_incidence_1k = mean(incidence_per_1k, na.rm = TRUE),
    .groups = "drop"
  )

landcover_malaria <- district_total_incidence %>%
  left_join(landcover_df, by = "district")

cat("Districts with land cover data:", sum(!is.na(landcover_malaria[[landcover_vars[1]]])), "\n")
head(landcover_malaria)
```

---

# Correlation Analysis

## Helper Functions

```{r helpers}
# Safe Pearson correlation returning a tibble
cor_test_safe <- function(x, y) {
  tryCatch({
    res <- cor.test(x, y, method = "pearson")
    tibble(correlation = unname(res$estimate), p_value = res$p.value)
  }, error = function(e) tibble(correlation = NA_real_, p_value = NA_real_))
}

# P-value to significance stars
p_to_stars <- function(p) {
  case_when(
    is.na(p)  ~ "",
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ ""
  )
}
```

## Land Cover vs Total Malaria Incidence per District

> **Note:** Land cover proportions are static (single snapshot per district), so
> they are correlated against incidence for a single representative year
> (**2020**) rather than the full 2015–2024 series, to avoid conflating a
> static covariate with a multi-year trend.

```{r landcover-correlation-total}
lc_cor_results <- map_dfr(landcover_vars, function(v) {
  cor_test_safe(landcover_malaria[[v]], landcover_malaria$total_incidence) %>%
    mutate(variable = v)
}) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ "ns"
    )
  ) %>%
  select(variable, correlation, p_value, significance)

kable(lc_cor_results,
      caption = "Pearson Correlations: Land Cover Class vs Total Malaria Incidence (2020)",
      digits  = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Land Cover vs Average Incidence per 1,000

```{r landcover-correlation-1k}
lc_cor_1k <- map_dfr(landcover_vars, function(v) {
  cor_test_safe(landcover_malaria[[v]], landcover_malaria$avg_incidence_1k) %>%
    mutate(variable = v)
}) %>%
  arrange(p_value) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ "ns"
    )
  ) %>%
  select(variable, correlation, p_value, significance)

kable(lc_cor_1k,
      caption = "Pearson Correlations: Land Cover Class vs Average Incidence per 1,000 (2020)",
      digits  = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Climate Variables vs Malaria Incidence (No Lag)

```{r climate-correlation-nolag}
climate_vars <- c("precipitation", "temperature", "wind")

combined_df <- merged_df %>%
  rename(`Malaria incidence` = incidence_per_1k) %>%
  mutate(across(everything(), ~ replace(., is.infinite(.), NA)))

# Safe correlation function (re-declared for this chunk)
cor_test_safe <- function(x, y) {
  tryCatch({
    res <- cor.test(x, y, method = "pearson")
    tibble(correlation = unname(res$estimate), p_value = res$p.value)
  }, error = function(e) tibble(correlation = NA_real_, p_value = NA_real_))
}

correlation_per_district <- combined_df %>%
  group_by(district) %>%
  summarise(
    across(
      all_of(climate_vars),
      ~ list(cor_test_safe(`Malaria incidence`, .x))
    ),
    .groups = "drop"
  )

cor_long <- correlation_per_district %>%
  pivot_longer(
    cols      = all_of(climate_vars),
    names_to  = "climate_variable",
    values_to = "result"
  ) %>%
  unnest(result) %>%
  mutate(stars = p_to_stars(p_value))
```

## Climate Variables vs Malaria Incidence (1-, 2-, and 3-Month Lags)

```{r climate-correlation-lag}
lagged_df <- combined_df %>%
  group_by(district) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(across(
    all_of(climate_vars),
    list(
      lag1 = ~ lag(.x, 1),
      lag2 = ~ lag(.x, 2),
      lag3 = ~ lag(.x, 3)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  ungroup()

climate_lag_vars <- grep(
  paste0("(", paste(climate_vars, collapse = "|"), ")_(lag1|lag2|lag3)"),
  names(lagged_df), value = TRUE
)

correlation_lags <- lagged_df %>%
  group_by(district) %>%
  summarise(
    across(
      all_of(climate_lag_vars),
      ~ list(cor_test_safe(`Malaria incidence`, .x)),
      .names = "{.col}"
    ),
    .groups = "drop"
  ) %>%
  mutate(across(all_of(climate_lag_vars), ~ map_dfr(.x, ~.x))) %>%
  unnest(cols = everything(), names_sep = "_")

cor_long_lags <- correlation_lags %>%
  pivot_longer(
    cols          = matches("_(lag[123])_"),
    names_to      = c("climate_variable", "lag", ".value"),
    names_pattern = "(.*)_(lag[123])_(.*)"
  ) %>%
  filter(!is.na(correlation)) %>%
  mutate(
    stars = p_to_stars(p_value),
    lag   = factor(lag,
                   levels = c("lag1", "lag2", "lag3"),
                   labels = c("1-Month Lag", "2-Month Lag", "3-Month Lag"))
  )

kable(cor_long_lags %>% arrange(district, climate_variable, lag),
      caption = "Climate–Malaria Correlations per District (1-, 2-, and 3-Month Lags)",
      digits  = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

# Visualization

## Malaria Incidence Over Time (2015–2024)

```{r viz-incidence-time, fig.width=16, fig.height=5}
malaria_monthly_max <- malaria_with_pop %>%
  group_by(date) %>%
  summarise(max_incidence = max(incidence, na.rm = TRUE), .groups = "drop")

full_dates <- data.frame(
  date = seq(as.Date("2015-01-01"), as.Date("2024-12-01"), by = "month")
)

malaria_monthly_max <- full_dates %>%
  left_join(malaria_monthly_max, by = "date") %>%
  mutate(month_label = format(date, "%b"))

year_start_dates <- malaria_monthly_max$date[format(malaria_monthly_max$date, "%m") == "01"]

ggplot(malaria_monthly_max, aes(x = date, y = max_incidence)) +
  geom_line(color = "darkred", linewidth = 1.2, na.rm = TRUE) +
  geom_point(color = "darkred", size = 1.8, na.rm = TRUE) +
  geom_vline(xintercept = as.numeric(year_start_dates),
             color = "grey50", linetype = "dashed") +
  scale_x_date(breaks = malaria_monthly_max$date,
               labels = malaria_monthly_max$month_label,
               limits = c(as.Date("2015-01-01"), as.Date("2024-12-01"))) +
  annotate("text", x = year_start_dates,
           y     = max(malaria_monthly_max$max_incidence, na.rm = TRUE) * 1.05,
           label = format(year_start_dates, "%Y"),
           vjust = 0, hjust = 0.5, size = 3.5) +
  expand_limits(y = max(malaria_monthly_max$max_incidence, na.rm = TRUE) * 1.1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 5)) +
  labs(title = NULL, x = NULL, y = "Incidence per 1,000")

ggsave("incidence_time_series.pdf", width = 16, height = 5, units = "in",
       device = "pdf")
```

```{r}
# Diagnostic: check which districts have missing population
district_monthly %>%
  filter(is.na(total_population) | total_population == 0) %>%
  distinct(district)

# Diagnostic: compare your rates vs expected
district_monthly %>%
  filter(year >= 2018) %>%
  group_by(district) %>%
  summarise(
    mean_pop      = mean(total_population, na.rm = TRUE),
    mean_cases    = mean(total_cases,      na.rm = TRUE),
    mean_per_1k   = mean(incidence_per_1k, na.rm = TRUE),
    n_missing_pop = sum(is.na(total_population)),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_per_1k)) %>%
  print(n = 30)
```


## Average Incidence per District (Ranked Bar Chart)

```{r viz-district-rank, fig.width=10, fig.height=7}
district_summary <- district_monthly %>%
  filter(year >= 2015, year <= 2024) %>%
  group_by(district) %>%
  summarise(
    mean_incidence_per_1k   = mean(incidence_per_1k,   na.rm = TRUE),
    median_incidence_per_1k = median(incidence_per_1k, na.rm = TRUE),
    max_incidence_per_1k    = max(incidence_per_1k,    na.rm = TRUE),
    total_cases             = sum(total_cases,          na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_incidence_per_1k))

print(district_summary)

ggplot(
  district_summary,
  aes(x = reorder(district, mean_incidence_per_1k), y = mean_incidence_per_1k)
) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = round(mean_incidence_per_1k, 1)),
    hjust = -0.1, size = 3
  ) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = NULL,
    x     = NULL,
    y     = "Mean Monthly Incidence per 1,000"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggsave("district_rank_bar.pdf", width = 10, height = 7, units = "in",
       device = "pdf")
```

## Monthly Incidence Heatmap per District

```{r viz-monthly-heatmap, fig.width=14, fig.height=10}
monthly_district_norm <- malaria_with_pop %>%
  mutate(month = factor(month, levels = month.name)) %>%
  group_by(district, month) %>%
  summarise(Incidence_per_1k = mean(incidence_per_1k, na.rm = TRUE), .groups = "drop")

ggplot(monthly_district_norm, aes(x = month, y = district, fill = Incidence_per_1k)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightyellow", high = "steelblue",
                      name = "Incidence\n(per 1,000)") +
  labs(title = NULL, x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    plot.title  = element_text(face = "bold", size = 14, hjust = 0.5)
  )

ggsave("monthly_incidence_heatmap.pdf", width = 14, height = 10, units = "in",
       device = "pdf")
```

## Annual Incidence per District Heatmap

```{r viz-annual-heatmap, fig.width=10, fig.height=7}
yearly_district <- malaria_with_pop %>%
  group_by(district, year) %>%
  summarise(Total_Incidence = sum(incidence_per_1k, na.rm = TRUE), .groups = "drop") %>%
  mutate(year = factor(year, levels = 2015:2024))

ggplot(yearly_district, aes(x = year, y = district, fill = Total_Incidence)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightyellow", high = "steelblue") +
  labs(title = NULL, x = NULL, y = NULL, fill = "Incidence") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title  = element_text(face = "bold", size = 10, hjust = 0.5)
  )

ggsave("annual_incidence_heatmap.pdf", width = 10, height = 7, units = "in",
       device = "pdf")
```

## Choropleth Maps of Malaria Incidence by Year

```{r viz-choropleth, fig.width=14, fig.height=10}
# Load shapefile
districts <- st_read("mwi_adm_nso_hotosm_20230405_shp/mwi_admbnda_adm2_nso_hotosm_20230405.shp")

# Verify
st_crs(districts)
print(names(districts))
print(head(districts$ADM2_EN))

# Check name alignment
cat("\nDistricts in shapefile but NOT in malaria data:\n")
print(setdiff(districts$ADM2_EN, unique(malaria_with_pop$district)))

cat("\nDistricts in malaria data but NOT in shapefile:\n")
print(setdiff(unique(malaria_with_pop$district), districts$ADM2_EN))

# Map
districts_sf_map <- st_as_sf(districts)

malaria_agg <- malaria_with_pop %>%
  filter(year >= 2015, year <= 2024) %>%
  group_by(district, year) %>%
  summarise(total_incidence = sum(incidence_per_1k, na.rm = TRUE), .groups = "drop")

malaria_map_data <- districts_sf_map %>%
  left_join(malaria_agg, by = c("ADM2_EN" = "district")) %>%
  filter(!is.na(total_incidence))

# Compute district centroids for labels — recode city entries to match district names
district_centroids <- districts_sf_map %>%
  st_centroid() %>%
  mutate(
    lon        = st_coordinates(.)[, 1],
    lat        = st_coordinates(.)[, 2],
    label_name = case_when(
      str_detect(tolower(ADM2_EN), "mzuzu")         ~ "Mzimba",
      str_detect(tolower(ADM2_EN), "lilongwe city") ~ "Lilongwe",
      TRUE ~ ADM2_EN
    )
  )

# Bounding box for arrow/scale placement
bbox <- st_bbox(districts_sf_map)

# Yearly choropleth — one facet per year, with North arrow and source caption
choropleth_map <- ggplot(malaria_map_data) +
  geom_sf(aes(fill = total_incidence), color = "black", linewidth = 0.2) +
  geom_sf_text(
    data = district_centroids,
    aes(label = label_name),
    size = 2.8,
    fontface = "bold",
    color = "black",
    check_overlap = TRUE
  ) +
  scale_fill_gradient(low = "#FFF7BC", high = "#D95F0E", name = "Incidence\nper 1,000") +
  facet_wrap(~year, ncol = 5) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    style = north_arrow_fancy_orienteering(
      fill      = c("black", "white"),
      text_size = 3
    ),
    height = unit(1.2, "cm"),
    width  = unit(1.2, "cm")
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.25,
    text_cex   = 0.75
  ) +
  theme_minimal(base_size = 11) +
  labs(
    title   = NULL,
    x       = NULL,
    y       = NULL,
    caption = "Source: Malawi shapefiles — Humanitarian Data Exchange (HDX) / HOT-OSM / NSO, 2023"
  ) +
  theme(
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    panel.grid      = element_blank(),
    strip.text      = element_text(face = "bold", size = 12),
    plot.caption    = element_text(size = 8, hjust = 0),
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 10)
  )

choropleth_map

# Save as PDF
ggsave(
  "choropleth_annual_malaria_incidence.pdf",
  plot = choropleth_map,
  width = 18, height = 9, units = "in", device = "pdf"
)
```

## Full Malawi Map (Descriptive Overview)

```{r viz-malawi-overview-map, fig.width=8, fig.height=12}
library(ggspatial)  # for north arrow and scale bar

# Step 1: Assign a single district name to every polygon, merging city polygons
#   into their parent district (Mzuzu City → Mzimba; Lilongwe City → Lilongwe)
districts_merged <- districts_sf_map %>%
  mutate(
    district_name = case_when(
      str_detect(tolower(ADM2_EN), "mzuzu")         ~ "Mzimba",
      str_detect(tolower(ADM2_EN), "lilongwe city") ~ "Lilongwe",
      TRUE ~ ADM2_EN
    )
  ) %>%
  group_by(district_name) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

# Step 2: One centroid per merged district → one label each
district_centroids_full <- districts_merged %>%
  st_centroid() %>%
  mutate(
    lon  = st_coordinates(.)[, 1],
    lat  = st_coordinates(.)[, 2],
    name = district_name
  )

# Malawi country bounding box approx dimensions:
# Width ~560 km (N–S ~900 km), Lake Malawi ~580 km long, ~50 km wide
malawi_overview <- ggplot(districts_sf_map) +
  geom_sf(fill = "#f5f0e8", color = "grey40", linewidth = 0.4) +
  geom_sf_text(
    data = district_centroids_full,
    aes(label = name),
    size  = 3.5,
    fontface = "bold",
    color = "black",
    check_overlap = TRUE
  ) +
  annotation_north_arrow(
    location    = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering(
      fill      = c("black", "white"),
      text_size = 10
    ),
    height = unit(1.5, "cm"),
    width  = unit(1.5, "cm")
  ) +
  annotation_scale(
    location   = "bl",
    width_hint = 0.3,
    text_cex   = 0.85
  ) +
  # Add Lake Malawi annotation (positioned at approximate lake centroid)
  annotate(
    "text",
    x = 34.5, y = -12.0,
    label = "Lake Malawi\n(~580 km long\n~50 km wide)",
    size = 3.2, color = "#2166ac", fontface = "italic",
    hjust = 0.5
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title   = NULL,
    x       = NULL,
    y       = NULL,
    caption = "Source: Malawi shapefiles — Humanitarian Data Exchange (HDX) / HOT-OSM / NSO, 2023"
  ) +
  theme(
    axis.text    = element_blank(),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank(),
    plot.caption = element_text(size = 8, hjust = 0)
  )

malawi_overview

# Save as PDF
ggsave(
  "malawi_districts_overview_map.pdf",
  plot = malawi_overview,
  width = 8, height = 12, units = "in", device = "pdf"
)
```

---

## Land Cover Correlation Plots

### Significance Bar Chart — Land Cover vs Total Incidence

```{r viz-lc-sig-bar, fig.width=10, fig.height=6}
lc_cor_results %>%
  filter(!is.na(correlation)) %>%
  mutate(
    variable  = fct_reorder(variable, correlation),
    variable  = factor(
      paste0(toupper(substr(as.character(variable), 1, 1)),
             substr(as.character(variable), 2, nchar(as.character(variable)))),
      levels = paste0(toupper(substr(levels(fct_reorder(variable, correlation)), 1, 1)),
                      substr(levels(fct_reorder(variable, correlation)), 2,
                             nchar(levels(fct_reorder(variable, correlation)))))
    ),
    direction = ifelse(correlation > 0, "Positive", "Negative")
  ) %>%
  ggplot(aes(x = variable, y = correlation, fill = direction)) +
  geom_col() +
  geom_text(
    aes(label = significance,
        y = correlation + ifelse(correlation >= 0, 0.03, -0.03)),
    size = 5, fontface = "bold"
  ) +
  geom_hline(yintercept = 0, linewidth = 0.6) +
  scale_fill_manual(values = c("Positive" = "#D7191C", "Negative" = "#2C7BB6")) +
  coord_flip() +
  labs(
    title    = NULL,
    subtitle = "Significance: *** p<0.001  ** p<0.01  * p<0.05  . p<0.1  ns",
    x        = NULL,
    y        = "Pearson r",
    fill     = "Direction"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_blank())

ggsave("lc_significance_bar.pdf", width = 10, height = 6, units = "in",
       device = "pdf")
```

### Significant Land Cover Classes — Ranked Dot Plot

```{r viz-lc-dotplot, fig.width=9, fig.height=5}
lc_cor_results %>%
  filter(significance != "ns", !is.na(correlation)) %>%
  mutate(variable = fct_reorder(variable, correlation)) %>%
  ggplot(aes(x = correlation, y = variable,
             color = significance, size = abs(correlation))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point() +
  geom_text(aes(label = significance), nudge_y = 0.35, size = 4.5, fontface = "bold") +
  scale_color_manual(values = c("***" = "#D7191C", "**" = "#E85D04",
                                 "*"  = "#F4A261", "."  = "#A8DADC")) +
  scale_size_continuous(range = c(3, 8), guide = "none") +
  labs(
    title    = NULL,
    subtitle = "Only p < 0.1 shown. Point size ∝ |r|.",
    x        = "Pearson r",
    y        = NULL,
    color    = "Significance"
  ) +
  theme_minimal(base_size = 12)

ggsave("lc_dotplot_significant.pdf", width = 9, height = 5, units = "in",
       device = "pdf")
```

### Land Cover vs Total Incidence — Scatter Plots

```{r viz-lc-scatter-total, fig.width=13, fig.height=10}
lc_sig_labels <- lc_cor_results %>%
  select(variable, significance, correlation) %>%
  rename(land_cover_class = variable)

# Compute per-facet regression equations and R²
lc_eqn_total <- map_dfr(landcover_vars, function(v) {
  df_sub <- landcover_malaria %>%
    select(cover_value = all_of(v), y = total_incidence) %>%
    filter(!is.na(cover_value), !is.na(y))
  if (nrow(df_sub) < 3) return(tibble(land_cover_class = v, eqn_label = NA_character_))
  fit    <- lm(y ~ cover_value, data = df_sub)
  cf     <- coef(fit)
  r2     <- summary(fit)$r.squared
  m      <- round(cf[2], 3)
  c_val  <- round(cf[1], 1)
  r2_val <- round(r2, 3)
  sign_c <- ifelse(c_val >= 0, "+", "-")
  label  <- paste0("y = ", m, "x ", sign_c, " ", abs(c_val),
                   "\nR² = ", r2_val)
  tibble(land_cover_class = v, eqn_label = label,
         # x/y position: top-left of facet
         x_pos = min(df_sub$cover_value, na.rm = TRUE),
         y_pos = max(df_sub$y,           na.rm = TRUE) * 0.95)
})

lc_long_total <- landcover_malaria %>%
  pivot_longer(cols = all_of(landcover_vars),
               names_to  = "land_cover_class",
               values_to = "cover_value") %>%
  left_join(lc_sig_labels, by = "land_cover_class") %>%
  mutate(
    # Capitalise first letter of land cover class name
    lc_label   = paste0(
      toupper(substr(land_cover_class, 1, 1)),
      substr(land_cover_class, 2, nchar(land_cover_class))
    ),
    facet_label = paste0(lc_label,
                         " (r=", round(correlation, 2), " ", significance, ")")
  )

lc_eqn_total <- lc_eqn_total %>%
  left_join(
    lc_long_total %>% select(land_cover_class, facet_label) %>% distinct(),
    by = "land_cover_class"
  )

ggplot(lc_long_total, aes(x = cover_value, y = total_incidence)) +
  geom_point(aes(color = significance), alpha = 0.8, size = 2.5) +
  geom_smooth(method = "lm", color = "black", se = TRUE, linewidth = 0.8) +
  geom_text_repel(aes(label = district), size = 2.5, max.overlaps = 6, color = "grey40") +
  geom_text(
    data = lc_eqn_total,
    aes(x = x_pos, y = y_pos, label = eqn_label),
    hjust = 0, vjust = 1, size = 2.8, color = "black", fontface = "italic",
    inherit.aes = FALSE
  ) +
  scale_color_manual(values = c("***" = "#D7191C", "**" = "#E85D04",
                                 "*"  = "#F4A261", "."  = "#A8DADC", "ns" = "grey70")) +
  facet_wrap(~facet_label, scales = "free_x") +
  labs(
    title    = NULL,
    subtitle = NULL,
    x        = "Land Cover Proportion",
    y        = "Total Malaria Incidence (2020)",
    color    = "Significance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text        = element_text(face = "bold", size = 9),
    strip.placement   = "outside",
    strip.background  = element_rect(color = "black", fill = "grey95", linewidth = 0.6),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.5),
    legend.position   = "bottom",
    axis.text         = element_text(size = 9)
  )

ggsave("lc_scatter_total_incidence.pdf", width = 13, height = 10, units = "in",
       device = "pdf")
```

### Land Cover vs Average Incidence per 1,000

```{r viz-lc-scatter-1k, fig.width=13, fig.height=10}
lc_sig_labels_1k <- lc_cor_1k %>%
  select(variable, significance, correlation) %>%
  rename(land_cover_class = variable)

lc_long_1k <- landcover_malaria %>%
  pivot_longer(cols = all_of(landcover_vars),
               names_to  = "land_cover_class",
               values_to = "cover_value") %>%
  left_join(lc_sig_labels_1k, by = "land_cover_class") %>%
  mutate(
    lc_label   = paste0(
      toupper(substr(land_cover_class, 1, 1)),
      substr(land_cover_class, 2, nchar(land_cover_class))
    ),
    facet_label = paste0(lc_label,
                         " (r=", round(correlation, 2), " ", significance, ")")
  )

ggplot(lc_long_1k, aes(x = cover_value, y = avg_incidence_1k)) +
  geom_smooth(method = "lm", color = "black", fill = "skyblue",
              alpha = 0.35, se = TRUE, linewidth = 0.8) +
  geom_point(aes(color = significance), alpha = 0.8, size = 2.5) +
  geom_text_repel(aes(label = district), size = 2.5, max.overlaps = 6, color = "grey40") +
  scale_color_manual(values = c("***" = "#D7191C", "**" = "#E85D04",
                                 "*"  = "#F4A261", "."  = "#A8DADC", "ns" = "grey30")) +
  facet_wrap(~facet_label, scales = "free_x") +
  labs(
    title    = NULL,
    subtitle = NULL,
    x        = "Land Cover Proportion",
    y        = "Average Incidence per 1,000 (2020)",
    color    = "Significance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text        = element_text(face = "bold", size = 9),
    strip.placement   = "outside",
    strip.background  = element_rect(color = "black", fill = "grey95", linewidth = 0.6),
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.5),
    legend.position   = "bottom",
    axis.text         = element_text(size = 9)
  )

ggsave("lc_scatter_avg_incidence_1k.pdf", width = 13, height = 10, units = "in",
       device = "pdf")
```

### Land Cover–Incidence Correlation Heatmap by District

```{r viz-lc-district-heatmap, fig.width=13, fig.height=9}
lc_dist_cor <- map_dfr(landcover_vars, function(v) {
  tibble(
    land_cover = v,
    district   = landcover_malaria$district,
    lc_value   = landcover_malaria[[v]],
    incidence  = landcover_malaria$total_incidence
  )
}) %>%
  filter(!is.na(lc_value), !is.na(incidence))

lc_dist_heat <- lc_dist_cor %>%
  left_join(
    lc_cor_results %>% rename(land_cover = variable),
    by = "land_cover"
  ) %>%
  mutate(stars = p_to_stars(p_value))

lc_dist_heat_plot <- lc_dist_heat %>%
  mutate(
    land_cover = paste0(
      toupper(substr(land_cover, 1, 1)),
      substr(land_cover, 2, nchar(land_cover))
    )
  )

ggplot(lc_dist_heat_plot, aes(x = land_cover, y = district, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = stars), color = "black", size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#2C7BB6", mid = "white", high = "#D7191C",
    midpoint = 0, limits = c(-1, 1), name = "Pearson r"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.title  = element_blank()
  ) +
  labs(
    title    = NULL,
    subtitle = NULL,
    x        = NULL,
    y        = NULL
  )

ggsave("lc_district_correlation_heatmap.pdf", width = 13, height = 9, units = "in",
       device = "pdf")
```

---

## Climate–Malaria Correlation Plots

### Correlation Heatmap — No Lag, per District

```{r viz-corr-nolag, fig.width=10, fig.height=9}
cor_long_plot <- cor_long %>%
  mutate(
    climate_variable = paste0(
      toupper(substr(climate_variable, 1, 1)),
      substr(climate_variable, 2, nchar(climate_variable))
    )
  )

ggplot(cor_long_plot, aes(x = climate_variable, y = district, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = stars), color = "black", size = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, name = "Correlation") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 10)
  ) +
  labs(
    title    = NULL,
    subtitle = "Significance: *** p<0.001  ** p<0.01  * p<0.05  . p<0.1",
    x        = NULL,
    y        = NULL
  )

ggsave("climate_correlation_nolag_heatmap.pdf", width = 10, height = 9, units = "in",
       device = "pdf")
```

### Correlation Heatmap — 1-, 2-, and 3-Month Lags, per District

```{r viz-corr-lag, fig.width=13, fig.height=9}
cor_long_lags_plot <- cor_long_lags %>%
  mutate(
    climate_variable = paste0(
      toupper(substr(climate_variable, 1, 1)),
      substr(climate_variable, 2, nchar(climate_variable))
    )
  )

ggplot(cor_long_lags_plot, aes(x = lag, y = climate_variable, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = stars), color = "black", size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1), name = "Correlation") +
  facet_wrap(~district, ncol = 4) +
  theme_minimal(base_size = 11) +
  labs(
    title    = NULL,
    subtitle = "Significance: *** p<0.001  ** p<0.01  * p<0.05  . p<0.1",
    x        = NULL,
    y        = NULL
  ) +
  theme(
    panel.grid  = element_blank(),
    strip.text  = element_text(face = "bold"),
    axis.text.x = element_text(size = 9)
  )

ggsave("climate_correlation_lagged_heatmap.pdf", width = 13, height = 9, units = "in",
       device = "pdf")
```

### Lag Comparison — Does Lagging Improve Correlations?

```{r viz-lag-comparison, fig.width=10, fig.height=5}
nolag_summary <- cor_long %>%
  group_by(climate_variable) %>%
  summarise(mean_r = mean(abs(correlation), na.rm = TRUE), .groups = "drop") %>%
  mutate(lag = "No Lag")

lag_summary <- cor_long_lags %>%
  group_by(climate_variable, lag) %>%
  summarise(mean_r = mean(abs(correlation), na.rm = TRUE), .groups = "drop") %>%
  mutate(lag = as.character(lag))

lag_compare <- bind_rows(nolag_summary, lag_summary) %>%
  mutate(lag = factor(lag, levels = c("No Lag", "1-Month Lag", "2-Month Lag", "3-Month Lag")))

lag_compare_plot <- lag_compare %>%
  mutate(
    climate_variable = paste0(
      toupper(substr(climate_variable, 1, 1)),
      substr(climate_variable, 2, nchar(climate_variable))
    )
  )

ggplot(lag_compare_plot, aes(x = lag, y = mean_r, fill = lag)) +
  geom_col(width = 0.6) +
  facet_wrap(~climate_variable) +
  scale_fill_manual(values = c("No Lag"      = "grey70",
                                "1-Month Lag" = "#64B5F6",
                                "2-Month Lag" = "#2196F3",
                                "3-Month Lag" = "#0D47A1")) +
  labs(
    title    = NULL,
    subtitle = "Averaged across all districts",
    x        = NULL,
    y        = "Mean |Pearson r|",
    fill     = "Lag"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text      = element_text(face = "bold"),
    legend.position = "none"
  )

ggsave("lag_comparison_bar.pdf", width = 10, height = 5, units = "in",
       device = "pdf")
```

### Climate vs Malaria Incidence — Scatter Plots (ERA5 Variables)

```{r viz-climate-scatter, fig.width=13, fig.height=9}
# Temperature
p_temp <- ggplot(combined_df, aes(x = temperature, y = `Malaria incidence`)) +
  geom_point(alpha = 0.3, color = "tomato") +
  geom_smooth(method = "lm", color = "darkred", se = TRUE) +
  facet_wrap(~district) +
  labs(title = NULL,
       x = "Temperature (°C)", y = "Incidence per 1,000") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 7))
p_temp
ggsave("scatter_temperature_vs_incidence.pdf", plot = p_temp,
       width = 13, height = 9, units = "in", device = "pdf")

# Precipitation
p_precip <- ggplot(combined_df, aes(x = precipitation, y = `Malaria incidence`)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "navy", se = TRUE) +
  facet_wrap(~district) +
  labs(title = NULL,
       x = "Precipitation (mm)", y = "Incidence per 1,000") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 7))
p_precip
ggsave("scatter_precipitation_vs_incidence.pdf", plot = p_precip,
       width = 13, height = 9, units = "in", device = "pdf")

# Wind
p_wind <- ggplot(combined_df, aes(x = wind, y = `Malaria incidence`)) +
  geom_point(alpha = 0.3, color = "darkseagreen") +
  geom_smooth(method = "lm", color = "darkgreen", se = TRUE) +
  facet_wrap(~district) +
  labs(title = NULL,
       x = "Wind Speed (m/s)", y = "Incidence per 1,000") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 7))
p_wind
ggsave("scatter_wind_vs_incidence.pdf", plot = p_wind,
       width = 13, height = 9, units = "in", device = "pdf")
```

---

# Final Export Dataset

Build a clean, analysis-ready dataset containing district-month incidence with same-month (no-lag) temperature and precipitation, plus 1-, 2-, and 3-month lagged rainfall and temperature. Wind is excluded from the export but kept in the source data.

```{r export-dataset}
# Build the lagged dataset from the merged_df (already has incidence_per_1k + climate)
export_df <- merged_df %>%
  arrange(district, date) %>%
  group_by(district) %>%
  mutate(
    # Rainfall lags (using raw precipitation)
    lag1_rainfall = lag(precipitation, 1),
    lag2_rainfall = lag(precipitation, 2),
    lag3_rainfall = lag(precipitation, 3),
    # Temperature lags
    lag1_temperature = lag(temperature, 1),
    lag2_temperature = lag(temperature, 2),
    lag3_temperature = lag(temperature, 3)
  ) %>%
  ungroup() %>%
  # Keep only rows where all lag columns are non-missing (i.e. lag-3 month is available)
  filter(!is.na(lag3_rainfall), !is.na(lag3_temperature)) %>%
  # Select and rename to the requested clean column set
  transmute(
    District               = district,
    Incidence_Month        = as.character(month),
    Year                   = year,
    Incidence_per_1k       = round(incidence_per_1k, 4),
    Temperature            = round(temperature,   3),
    Precipitation          = round(precipitation, 2),
    lag1_rainfall          = round(lag1_rainfall,    2),
    lag2_rainfall          = round(lag2_rainfall,    2),
    lag3_rainfall          = round(lag3_rainfall,    2),
    lag1_temperature       = round(lag1_temperature, 3),
    lag2_temperature       = round(lag2_temperature, 3),
    lag3_temperature       = round(lag3_temperature, 3)
  ) %>%
  # Reorder columns to the requested layout
  select(
    District,
    Incidence_Month,
    Year,
    Incidence_per_1k,
    Temperature,
    Precipitation,
    lag1_rainfall,
    lag2_rainfall,
    lag3_rainfall,
    lag1_temperature,
    lag2_temperature,
    lag3_temperature
  ) %>%
  arrange(District, Year, match(Incidence_Month, month.name))

cat("Rows in export dataset:", nrow(export_df), "\n")
cat("Columns:", paste(names(export_df), collapse = ", "), "\n")
glimpse(export_df)

# ── Export as CSV ────────────────────────────────────────────────────────────
write_csv(export_df, "malaria_lagged_climate_export.csv")

# ── Export as Excel (formatted) ──────────────────────────────────────────────
wb <- createWorkbook()
addWorksheet(wb, "Malaria_Lagged_Climate")

# Header style
header_style <- createStyle(
  fontColour = "#FFFFFF",
  fgFill     = "#2C5F8A",
  halign     = "CENTER",
  textDecoration = "bold",
  border     = "Bottom"
)

# Write data
writeData(wb, sheet = 1, export_df, startRow = 1, startCol = 1)

# Apply header style
addStyle(wb, sheet = 1, style = header_style,
         rows = 1, cols = 1:ncol(export_df), gridExpand = TRUE)

# Auto-width columns
setColWidths(wb, sheet = 1, cols = 1:ncol(export_df), widths = "auto")

# Freeze top row
freezePane(wb, sheet = 1, firstRow = TRUE)

# Alternating row shading
row_style <- createStyle(fgFill = "#EBF3FB")
for (r in seq(2, nrow(export_df) + 1, by = 2)) {
  addStyle(wb, sheet = 1, style = row_style,
           rows = r, cols = 1:ncol(export_df), gridExpand = TRUE)
}

saveWorkbook(wb, "malaria_lagged_climate_export.xlsx", overwrite = TRUE)

message("Export complete: malaria_lagged_climate_export.csv and .xlsx")
```

---

# Interpretation

## Land Cover and Malaria Incidence

The two correlation tables (total incidence and per-1,000 incidence) rank each land cover class by its association with malaria burden across Malawi's districts. Classes with significant **positive** correlations likely represent environments that support mosquito breeding — such as wetlands, irrigated cropland, or dense shrubland near water bodies. Classes with significant **negative** correlations (e.g., urban surfaces, bare land) typically reflect environments less conducive to *Anopheles* breeding due to impervious surfaces or reduced moisture retention.

The regression equations (y = mx + c) displayed on each scatter facet quantify the linear relationship: **m** is the gradient (change in incidence per unit of land cover) and **c** is the intercept. The R² values indicate the proportion of variance in incidence explained by each cover class alone. High |m| with high R² signals both a strong and precise predictor.

## Climate Variables and Malaria Incidence (No Lag)

The contemporaneous correlation heatmap shows within-month relationships between ERA5 variables and incidence. Given the *Anopheles* life cycle (≈2–4 weeks from egg to infectious bite), same-month correlations are expected to be modest. Districts showing strong same-month temperature correlations may lie in regions where temperatures are already near the optimal range for parasite development year-round.

## Lagged Climate Correlations (1-, 2-, and 3-Month Lags)

The 1-, 2-, and 3-month lag analysis is central to predictive modelling. **Precipitation** is expected to show its strongest positive correlation at lag 1–3 months, reflecting the time needed for standing water to accumulate and larval cohorts to reach adulthood. **Temperature** typically acts at shorter lags because it primarily modulates the extrinsic incubation period of *Plasmodium falciparum* (≈10–14 days at 25°C). **Wind** is generally a weak predictor.

## Summary for Model Development

| Variable          | Expected Lag | Rationale                                       |
|-------------------|-------------|--------------------------------------------------|
| Precipitation     | 1–3 months  | Breeding site formation and larval development   |
| Temperature       | 0–1 month   | Affects extrinsic incubation period              |
| Wind speed        | 0 months    | Weak causal pathway; include as control only     |
| Land cover class  | Static      | Background susceptibility modifier per district  |

Variables showing consistent, statistically significant correlations at a given lag should be prioritised as features in the subsequent predictive model (e.g., negative binomial regression, random forest, or LSTM). Land cover variables should be included as static district-level covariates. The exported dataset (`malaria_lagged_climate_export.csv` / `.xlsx`) is ready for direct model ingestion.
