---
title: "Baseline Models"
output: html_document
author: "Victor Arraes Rocha Felix"
date: "2026-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(forecast)
library(ggplot2)
library(zoo)
library(tidyverse)
library(ggplot2)
library(lubridate) #dealing with dates
library(broom)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(patchwork)
library(tseries)
library(lme4)
library(forecast)
library(kableExtra)
library(tidyr)
library(tibble)
```

Loading auxiliary functions

```{r}
source("../aux_functions.R")
```

Let's start loading the data district name, Malaria incidence and lagged covariates

```{r}
correct_incidence_data <- read_csv("../malaria_lagged_climate_export.csv")

head(correct_incidence_data)
```

Split the data by district creating a list of dataframes and make sure the data is in a correct order by Date.

```{r}

# Transforming the data in a list of dataframes by District
district_data_list <- correct_incidence_data %>%
  mutate(
    Date = my(paste(Incidence_Month, Year))
  ) %>%
  arrange(District, Date) %>%
  group_by(District) %>%
  group_split() %>%
  setNames(unique(sort(correct_incidence_data$District)))

```

Fit the baseline of each district (single dataframe in the list) on a rollinnf window of lengh 60 which makes up to 22 windows across the whole data.

```{r}

# Fit each model for each district on a rolling window
baseline_models <- lapply(
  district_data_list,
  rolling_baseline,
  n_windows = 22,
  window_size = 60
)

```

Forecast 3 months ahead (h=3) for each fitted model.

```{r}

# Baseline forecasts
BASELINE_model_forecasts <- lapply(
  baseline_models,
  forecast_baselines1,
  h = 3
)

```

Get the forecasted results as quantiles and also get the forecast date.

```{r}

# Get quantiles and forecats date
BASELINE <- lapply(
  BASELINE_model_forecasts,
  get_forecasted_quantiles1
)

```

Formatting the forecasted data so we can later calculate WIS and MAE.

```{r}

#formatted_baseline_forecasts <- data.frame()

#for (district_name in names(BASELINE)) {

#  forecast_of_one_district <- format_forecast_for_WIS(
#    BASELINE[[district_name]]
#  )

#  forecast_of_one_district$location <- district_name
#  forecast_of_one_district$model <- "BASELINE"

#  formatted_baseline_forecasts <- bind_rows(
#    formatted_baseline_forecasts,
#    forecast_of_one_district
#  )
#}


formatted_baseline_forecasts<-format_district_forecasts(BASELINE, model_name = "BASELINE")

# take a look at the data
head(formatted_baseline_forecasts)

```

Set all values bellow 0 to 0, to keep only possible forecasted values.

```{r}

formatted_baseline_forecasts$value[
  formatted_baseline_forecasts$value < 0
] <- 0

```

Save baseline forecasts

```{r}

write.csv(formatted_baseline_forecasts, "baseline_forecasts.csv", row.names = FALSE)

```


Formatting the initial data so we can compare the forecasts with actual values.

```{r}

head(correct_incidence_data)

truth <- correct_incidence_data %>%
  mutate(
    target_end_date = my(paste(Incidence_Month, Year))
  ) %>%
  transmute(
    target_end_date,
    location = District,
    value = Incidence_per_1k,
    target_variable = "incidence",
    model = "BASELINE"
  )

head(truth)

```

Get the Weighted Interval Scores (WIS) and Absolute Error of each forecast based on actual values.

```{r}

BASELINE_WIS_and_MAE<-covidHubUtils::score_forecasts(formatted_baseline_forecasts, truth)

write.csv(BASELINE_WIS_and_MAE, "baseline_wis_and_mae.csv", row.names = FALSE)

```

Plot all forecasts and the actual data

```{r}

forecast_plot_data <- formatted_baseline_forecasts %>%
  filter(
    model == "BASELINE",
    type == "quantile",
    quantile %in% c(0.025, 0.5, 0.975)
  ) %>%
  tidyr::pivot_wider(
    names_from = quantile,
    values_from = value
  )

plots <- forecast_plot_data %>%
  split(.$location) %>%
  lapply(function(df) {
    
    loc <- unique(df$location)
    
    truth_loc <- truth %>%
      filter(location == loc)
    
    ggplot(df, aes(
      x = target_end_date,
      group = forecast_date
    )) +
      geom_ribbon(aes(
        ymin = `0.025`,
        ymax = `0.975`,
        fill = as.factor(forecast_date)
      ), alpha = 0.2, color = NA) +
      geom_line(aes(
        y = `0.5`,
        color = as.factor(forecast_date)
      )) +
      geom_point(aes(
        y = `0.5`,
        color = as.factor(forecast_date)
      )) +
      geom_line(
        data = truth_loc,
        aes(
          x = target_end_date,
          y = value
        ),
        inherit.aes = FALSE,
        color = "black",
        linewidth = 1
      ) +
      geom_point(
        data = truth_loc,
        aes(
          x = target_end_date,
          y = value
        ),
        inherit.aes = FALSE,
        color = "black",
        size = 2
      ) +
      theme_minimal() +
      labs(
        x = "Target end date",
        y = "Incidence per 1k",
        color = "Forecast date",
        fill = "Forecast date",
        title = loc
      )+ theme(legend.position = "none")
  })

plots
```