---
title: "SARIMA Models"
output: html_document
date: "2026-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(forecast)
library(ggplot2)
library(zoo)
library(tidyverse)
library(ggplot2)
library(lubridate) #dealing with dates
library(broom)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(patchwork)
library(tseries)
library(lme4)
library(forecast)
library(kableExtra)
library(tidyr)
library(tibble)
```

Loading auxiliary functions

```{r}
source("aux_functions.R")
```

Let's start loading the data district name, Malaria incidence and lagged covariates

```{r}
correct_incidence_data <- read_csv("malaria_lagged_climate_export.csv")

head(correct_incidence_data)
```

Split the data by district creating a list of dataframes and make sure the data is in a correct order by Date.

```{r}

# Transforming the data in a list of dataframes by District
district_data_list <- correct_incidence_data %>%
  mutate(
    Date = my(paste(Incidence_Month, Year))
  ) %>%
  arrange(District, Date) %>%
  group_by(District) %>%
  group_split() %>%
  setNames(unique(sort(correct_incidence_data$District)))

```

Fit the baseline of each district (single dataframe in the list) on a rollinnf window of lengh 60 which makes up to 22 windows across the whole data.

```{r}

# Fit each model for each district on a rolling window
sarima_models <- lapply(
  district_data_list,
  rolling_sarima,
  n_windows = 22,
  window_size = 60
)

```

Forecast 3 months ahead (h=3) for each fitted model.

```{r}

# Baseline forecasts
SARIMA_model_forecasts <- lapply(
  sarima_models,
  forecast_SARIMAs1,
  h = 3
)

```

Get the forecasted results as quantiles and also get the forecast date.

```{r}

# Get quantiles and forecats date
SARIMA <- lapply(
  SARIMA_model_forecasts,
  get_forecasted_quantiles1
)

```

Formatting the forecasted data so we can later calculate WIS and MAE.


```{r}

formatted_SARIMA_forecasts <- format_district_forecasts(
  SARIMA,
  model_name = "SARIMA"
)


# take a look at the data
head(formatted_SARIMA_forecasts)
```

Set all values bellow 0 to 0, to keep only possible forecasted values.

```{r}

formatted_SARIMA_forecasts$value[
  formatted_SARIMA_forecasts$value < 0
] <- 0

```

Save baseline forecasts

```{r}

write.csv(formatted_SARIMA_forecasts, "SARIMA_forecasts.csv", row.names = FALSE)

```


Formatting the initial data so we can compare the forecasts with actual values.

```{r}

truth <- correct_incidence_data %>%
  mutate(
    target_end_date = my(paste(Incidence_Month, Year))
  ) %>%
  transmute(
    target_end_date,
    location = District,
    value = Incidence_per_1k,
    target_variable = "incidence",
    model = "SARIMA"
  )

head(truth)

```

Get the Weighted Interval Scores (WIS) and Absolute Error of each forecast based on actual values.

```{r}

SARIMA_WIS_and_MAE<-covidHubUtils::score_forecasts(formatted_SARIMA_forecasts, truth)

write.csv(SARIMA_WIS_and_MAE, "sarima_wis_and_mae.csv", row.names = FALSE)

```